---
title: "Betting Against Beta"
author: "Silverio J. Vasquez"
date: "December 8, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(RSQLite)
require(quantmod)
require(Quandl)
require(tidyr)
require(dplyr)
require(PerformanceAnalytics)
require(rvest)
require(stringr)
require(readr)
require(data.table)
require(ggplot2)
require(statar)
require(curl)
setwd("~/Documents/bab")
```

# Betting Against Beta in the S&P 500

Below is my final project for Data 607. Some new features used that weren't covered in class were `SQLite`, `data.table`, and `PerformanceAnalytics`. First time this code is executed, it takes about an hour from start to finish running on an average Ubuntu computer (a bit longer on an average Windows PC). Subsequent runs are seamless once betas and prices are saved in SQLite. 

**BACKGROUND:**

In most standard Finance 101 courses, students are taught that the higher the risk, the higher the reward (i.e., only take on more risk, if reward increases as well). This ideology is ingrained in many theories, most notably in the Capital Asset Pricing Model where the return of a stock is linearly associated to the risk sensitivity (i.e., beta) to the broader equity market (e.g., S&P 500). As a result, individuals that aren't risk averse and wanted higher returns, ended up taking on more risk in their portfolio by selecting risker (higher beta) stocks.

However, in 2010, Andrea Frazzini and Lasse H. Pedersen published a paper titled "Betting Against Beta" (http://www.nber.org/papers/w16601) where they show empirically that lower beta stocks (less risky stocks) actually outperform higher beta stocks (higher risky stocks). 

Many investors have been able to reproduce similar results and some have even launched investment funds following this methodology. 

## DATA

**Sources:**
The data are sourced from [Wikipedia](https://en.wikipedia.org/wiki/List_of_S%26P_500_companies), [Fama and French's website](http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html), Quandl.com, and the `quantmod` R package.

**There is a large zip file with daily prices for a large number of stocks downloaded from [Quandl.com](https://www.Quandl.com). The only way to ensure 100% complete reproducibility of this project is to make this large zip file available via my [Dropbox](https://www.dropbox.com/sh/fzih0uypabr2a99/AAA_i1aa5XG1NpAQIttH2Tdna?dl=0). This file needs to be downloaded and extracted. The CSV that will be extracted is about 1.8gbs. This CSV should be extracted in the folder you'll ultimately set as your working directory and rename the CSV file to "WIKI_PRICES_ALL.csv".** This working directory is also where the SQLite database will be saved with all necessary data to have subsequent analysis run more efficiently.

Initially, the data will be sourced from the various sources mentioned above. This will cause some parts of the code to run for 20mins. In total, the entire script should take about 40mins to an hour first time around. Since this takes considerable time, the necessary dataframes will be saved in a SQLite database in table called "weeky_prices". This way every subsequent run of the `R` code will run smoothly.

```{r data}

# --------------------------------------------------------------
# Read Fama French data from CSV file
# --------------------------------------------------------------
ff <- read_csv(curl("https://raw.githubusercontent.com/sjv1030/Data607-Final/master/F-F_Research_Data_5_Factors_2x3_daily.CSV"),skip=3)
colnames(ff) <- c("Date","Mkt-RF","SMB","HML","CMA","risk.free")
ff$Date <- as.Date.character(ff$Date,format = "%Y%m%d")
ff <- as.xts(ff[,-1],order.by = ff$Date)
wk.end <- endpoints(ff,on="weeks")
ff_w <- ff[wk.end,]
plot(ff_w$risk.free,main="Risk-Free Rate")
# --------------------------------------------------------------

# --------------------------------------------------------------
# Check if "weekly_prices" table exists in database
# If not, then scrape Wikipedia page for S&P 500 tickers
# And download data from Yahoo Finance via quantmod
# Then save weekly prices to database
# --------------------------------------------------------------

# Scrape S&P500 company tickers from Wikipedia
url <- 'https://en.wikipedia.org/wiki/List_of_S%26P_500_companies'
wiki <- read_html(url) %>% html_nodes('table') %>% html_table(fill = TRUE)
names(wiki[[1]])
sp500 <- wiki[[1]] %>% select(1,2,4,5)

names(wiki[[2]])
add_memb <- wiki[[2]][-(0:2),] %>% select(2,3,4,5)

tmp <- c(sp500[,1],add_memb$X2,add_memb$X4)

print(head(tmp))

# Create date index
idx <- seq(from=as.Date('2000-01-13'),to=as.Date('2017-09-07'),by="weeks")

db <- dbConnect(SQLite(), dbname="betas.sqlite")  
if (dbExistsTable(db, "weekly_prices")) {
  # If weekly_prices table exists, pull in data from db
  df_all <- dbGetQuery(conn=db, "SELECT * FROM weekly_prices")
  df_all <- as.xts(df_all, order.by=idx)
  dbDisconnect(db)
} else {

# Progress bar
n <- length(tmp)
pb <- txtProgressBar(min = 0, max = n, style=3)

dataset <- xts() 
for(i in 1:length(tmp)) {
  if (tmp[i] == "") next # skip to next iteration
  symbol <- tmp[i]
  tryit <- try(getSymbols(symbol, from="2000-01-01", to="2017-09-30", src='yahoo'))
  if(inherits(tryit, "try-error")){
    i <- i+1
  } 
  else {
    data <- getSymbols(symbol, from="2000-01-01", to="2017-09-30", src='yahoo')
    dataset <- merge(dataset, Ad(get(tmp[i])))
    rm(symbol)
  }
  setTxtProgressBar(pb, i)
}

# Convert dataset to weekly
wk.end <- endpoints(dataset,on="weeks")
data_m <- dataset[wk.end,]
rets <- diff(log(data_m))[-1,]

# Get S&P 500 (SPX) returns
getSymbols("^GSPC", from="2000-01-01", to="2017-09-30", src="yahoo")
spx <- Ad(GSPC)
names(spx) <- "spx"
wk.end <- endpoints(spx,on="weeks")
spx_m <- spx[wk.end,]
ret_spx <- diff(log(spx_m))[-1,]

# Replace column names
headings <- colnames(rets) %>% str_replace_all(".Adjusted","")
names(rets) <- headings

# Identify stocks not downloaded
filter <- tmp %in% colnames(rets)
tmp2 <- tmp[!filter]

# Use power of Data Table to read in large CSV file 
# File has closing prices of LOTS of stocks
# Then filter data table for stocks that were not downloaded
# And convert from long to wide form
system.time(w <- fread("WIKI_PRICES_ALL.csv"))
w2 <- w[ticker %in% tmp2]
other_stks <- w2 %>% 
  select(date,ticker,adj_close) %>% 
  spread(ticker, adj_close)

# Convert dataset to weekly
other_stks$date <- as.Date(other_stks$date, "%Y-%m-%d")
other_stks.ts <- as.xts.data.table(other_stks)
wk.end <- endpoints(other_stks,on="weeks")
other_stks.w <- other_stks.ts[wk.end,]
other_rets <- diff(log(other_stks.w))[-1,]

# Merge returns with other_stks data frame
rets2 <- merge.xts(rets,other_rets,join="inner")

# Merge returns, SPX, and risk free rate from Fama French dataset
rets3 <- merge.xts(rets2,ret_spx,join="inner")
df_all <- merge.xts(rets3,ff_w$risk.free, join="inner")

# Store weekly prices in database
df_all_ <- as.data.frame(df_all)
dbWriteTable(conn = db, name="weekly_prices", value=df_all_, overwrite=TRUE)

# Disconnect database
dbDisconnect(db)
}

# Plot a few of the stocks and return series to confirm data download
draw <- function(x,i){
  p <- qplot(x=index(x),y=x[,i], geom="line", xlab="",ylab="",main=colnames(x)[i])
  print(p)
}

t <- sample(1:ncol(df_all),10,replace=FALSE)
for (i in t){ draw(df_all,i)}
```

## METHODOLOGY

**Calculate Beta:**

The approach to calculating beta below differs from the one in the research paper. The methodolgy taken here is to calculate the beta as the coefficient from a rolling 2-year regression of returns less the risk-free rate using the CAPM formula. Then Blume's adjustment is applied on beta, which is the concept that long-term betas are roughly 1.

The regression equation is below, which is a simple reaggrangement of the CAPM formula.

**CAPM Formula**

$$ Return = Riskfree Rate + \beta * (Market Return - Riskfree Rate) $$

$$ (Return - Riskfree Rate) = \beta * (Market Return - Riskfree Rate) $$
**Blume's Beta Adjustment**

$$ \bar\beta = \frac{2}{3} * \beta + \frac{1}{3} * 1 $$

**Got Beta?**

If betas aren't in the database initially, then a FOR loop creates a dataframe of rolling regression outcomes. Only companies with at least 3-years worth of weekly returns are considered. The regression is without an intercept as illustrated in the formula above. The beta dataframe is then saved to the SQLite database in table "weekly_betas".

Note: the FOR loop takes about 20mins to run.

```{r beta}
# --------------------------------------------------------------
# Check if weekly_betas exist in database
# If not, then calculate betas and save to database
# --------------------------------------------------------------
db <- dbConnect(SQLite(), dbname="betas.sqlite")  
if (dbExistsTable(db, "weekly_betas")) {
  # If betas database exists, pull in data this way
  betas <- dbGetQuery(conn=db, "SELECT * FROM weekly_betas")
  betas <- as.xts(betas,order.by = idx)
  dbDisconnect(db)
} else {
n <- ncol(df_all)-2
betas <- data.frame()

# Progress bar
pb <- txtProgressBar(min = 0, max = n, style=3)

for (i in 1:n){
  if(all(is.na(df_all[,i]))) next # skip to next iteration
  tmp_df <- na.omit(cbind(df_all[,i],df_all$spx,df_all$risk.free))
  tmp_df$exrt <- tmp_df[,1] - tmp_df$risk.free
  tmp_df$spx.rf <- tmp_df$spx - tmp_df$risk.free

  if(nrow(tmp_df)<52*3) next # skip to next iteration

  rollingbeta <- rollapply(tmp_df,
                           width=52*2,
                           FUN = function(Z)
                           {
                             t = lm(formula=exrt~spx.rf-1, data = as.data.frame(Z), na.action=na.omit);
                             return(t$coef*2/3+1/3)
                           },
                           by.column=FALSE, align="right")

  col_name <- names(df_all)[i]
  names(rollingbeta) <- col_name
  betas <- merge.xts(betas,rollingbeta)
  setTxtProgressBar(pb, i)
}

betas_ <- as.data.frame(betas)
dbWriteTable(conn = db, name="weekly_betas", value=betas_, overwrite=TRUE)

dbDisconnect(db)
}
# --------------------------------------------------------------
```

**Companies Not Included:**
```{r not included}
# --------------------------------------------------------------
# Create logic vector to see which names companies
# didn't have enough data to calculate a beta for
# And print tickers of those companies
in_in <- names(df_all) %in% names(betas)
print(names(df_all[,!in_in]))
# --------------------------------------------------------------
```

## RESULTS

**Weekly Rebalancing:**

A function called "getSig" is created, which takes a row of betas and buckets them into deciles on a weekly basis. The top 30% of betas are selected in the high_beta dataframe (stocks to be sold short) and low_beta dtaframes (stocks to be bought). 

The dataframe of +1s and 0s is then multiplied to the matrix of betas. All betas receive a weight proportional to the size of the beta for each week (i.e., row). This implies weekly rebalancing of weights for stocks being traded.

The returns are then aggregated into two portfolios: one that is long low-beta stocks and another that is short high-beta stocks. These two portfolios are then added together to make an equal-weighted total portfolio return (i.e., the total return an investor will receive from investing in both long and short portfolios equally).


```{r signal}
# Winsorize the betas dataset to handle outliers
betas2 <- t(apply(betas,1,function(x) winsorise(x,probs = c(0.05,0.95),verbose = FALSE)))

# Create two different datasets for betas above 1 and below 1
# The result is a dataframe of beta differences from 1
high_beta <- ifelse(betas2>1,betas2-1,0)
low_beta <- ifelse(betas2<1,1-betas2,0)

# Get signal selects top 30% of names by giving it a 1 and 0 for anything else
getSig <- function(row) {
  q <- quantile(row,probs=seq(0,1,0.1),na.rm=TRUE)
  h <- ifelse(row > q[8],1,0)
}

# The getSig function is then applied to the high and low beta dataframes
# resulting in dataframes of 1s & 0s that will serve to select the stocks
# that will ultimately be traded
high_sig <- t(apply(high_beta,1,getSig))
low_sig <- t(apply(low_beta,1,getSig))

high_stks <- as.matrix(high_beta) * high_sig
low_stks <- as.matrix(low_beta) * low_sig

# For each row (each week of betas), weights are created depending on
# how far away the beta is from 1
# The weights for the dataframe of stocks to be sold short are multiplied by -1
short_wt <- lag(high_stks/rowSums(high_stks, na.rm = TRUE),1)[-1,]*-1
long_wt <- lag(low_stks/rowSums(low_stks,na.rm=TRUE),1)[-1,]

# The following set of code confirms that the weights for each column in total equal 1 (i.e., 100%) for each dataframe of weights
chk_sh_wt <- apply(short_wt,1,function(x) sum(x,na.rm=TRUE))
plot(chk_sh_wt,type="b",main="Total Weight of Stocks in Short Portfolio")
chk_lo_wt <- apply(long_wt,1,function(x) sum(x,na.rm=TRUE))
plot(chk_lo_wt,type="b",main="Total Weight of Stocks in Long Portfolio")

# The following set of code shows how many stocks (roughly) will be traded
# for each long and short side
sh_vec <- apply(short_wt,1,function(x) sum(x<0,na.rm=TRUE))
plot(sh_vec,type="l", main="Number of Stocks in Short Portfolio")
lg_vec <- apply(long_wt,1,function(x) sum(x>0,na.rm=TRUE))
plot(lg_vec,type="l",main="Number of Stocks in Long Portfolio")

# There's no need for NAs anymore in the weight dataframes
# All NAs are converted to 0s since an NA is essentially the same as not giving
# a stock any weight in the portfolio
# This is done to calculate portfolio returns
long_wt[is.na(long_wt)] <- 0
short_wt[is.na(short_wt)] <- 0

# Subset returns dataframe to only have returns for stocks whose betas were calculated
# First row is removed to match dimension of weight dataframes
stks <- df_all[,in_in][-1,] 
stks[is.na(stks)] <- 0 # Convert NAs to 0s

# Pret_long is the portfolio return of the stocks that were bought
# Pret_short is the portfolio return of the stocks that were sold short
pret_long <- long_wt * stks
pret_short <- short_wt * stks

# P is the sum (really weighted sum given returns were multiplied by weights above) of all the returns of stocks that were bought on a weekly 
# Same is true for S for the stocks that were sold short on a weekly basis
p <- apply(pret_long,1,sum) 
s <- apply(pret_short,1,sum)
pt <- cbind.data.frame(p,s) # combine both p & s into a dataframe
idx <- seq(as.Date("2000-01-20"),as.Date("2017-09-07"),by = "week")
pt <- as.xts(pt,order.by = idx) # convert to xts object
f <- p + s # calculate total portfolio return if long and short portfolios were equal weighted
f.ts <- as.xts(f,order.by = idx) # convert total portfolio return to xts object

# Cumulative return is then calculated and plotted. Anything below the horizontal line means the investment lost money. Other performance statistics are then plotted using functions from the PerformanceAnalytics package.
cumret <- cumprod(1+f.ts)
plot.xts(cumret)
abline(h=1)
charts.PerformanceSummary(f.ts)
Return.annualized(f.ts)
SharpeRatio.annualized(f.ts)
```

**General Rebalancing:**

The code above is then modified slightly so that it can be called as a function taking in only one paramater - frequency of rebalance. The options are weekly (default setting), monthly, quarterly, or annually.

```{r general-rebal}
bab_test <- function(x="week"){
  t_endpoints <- endpoints(betas,x)
  betas_ <- betas[t_endpoints,]
  betas2 <- t(apply(betas_,1,function(x) winsorise(x,probs = c(0.05,0.95),verbose = FALSE)))

  # Create two different datasets for betas above 1 and below 1
  # The result is a dataframe of beta differences from 1
  high_beta <- ifelse(betas2>1,betas2-1,0)
  low_beta <- ifelse(betas2<1,1-betas2,0)
  
  # Get signal selects top 30% of names by giving it a 1 and 0 for anything else
  getSig <- function(row) {
    q <- quantile(row,probs=seq(0,1,0.1),na.rm=TRUE)
    h <- ifelse(row > q[8],1,0)
  }
  
  # The getSig function is then applied to the high and low beta dataframes
  # resulting in dataframes of 1s & 0s that will serve to select the stocks
  # that will ultimately be traded
  high_sig <- t(apply(high_beta,1,getSig))
  low_sig <- t(apply(low_beta,1,getSig))
  
  high_stks <- as.matrix(high_beta) * high_sig
  low_stks <- as.matrix(low_beta) * low_sig
  
  # For each row (each week of betas), weights are created depending on
  # how far away the beta is from 1
  # The weights for the dataframe of stocks to be sold short are multiplied by -1
  short_wt <- lag(high_stks/rowSums(high_stks, na.rm = TRUE),1)[-1,]*-1
  long_wt <- lag(low_stks/rowSums(low_stks,na.rm=TRUE),1)[-1,]
  
  # The following set of code confirms that the weights for each column in total equal 1 (i.e., 100%) for each dataframe of weights. Weights are 0 in beginning when there aren't any betas.
  chk_sh_wt <- apply(short_wt,1,function(x) sum(x,na.rm=TRUE))
  plot(chk_sh_wt,type="b",main="Total Weight of Stocks in Short Portfolio")
  chk_lo_wt <- apply(long_wt,1,function(x) sum(x,na.rm=TRUE))
  plot(chk_lo_wt,type="b",main="Total Weight of Stocks in Long Portfolio")
  
  # There's no need for NAs anymore in the weight dataframes
  # All NAs are converted to 0s since an NA is essentially the same as not giving
  # a stock any weight in the portfolio
  # This is done to calculate portfolio returns
  long_wt[is.na(long_wt)] <- 0
  short_wt[is.na(short_wt)] <- 0
  if (x == "week" || x == "months"){
    idx <- seq(as.Date("2000-01-20"),as.Date("2017-09-07"),by = x)
  } else if(x == "quarter") {
    idx <- seq(as.Date("2000-06-01"),as.Date("2017-09-01"),by = x)
  } else {
    idx <- seq(as.Date("2001-01-01"),as.Date("2017-09-01"),by = x)
  }
  long_wt <- as.xts(long_wt,order.by=idx)
  short_wt <- as.xts(short_wt,order.by=idx)
  
  long_wt <- long_wt['2002-02/2017-06']
  short_wt <- short_wt['2002-02/2017-06']
  
  # The following set of code shows how many stocks (roughly) will be traded
  # for each long and short side.
  sh_vec <- apply(short_wt,1,function(x) sum(x<0,na.rm=TRUE))
  plot(sh_vec,type="l", main="Number of Stocks in Short Portfolio")
  lg_vec <- apply(long_wt,1,function(x) sum(x>0,na.rm=TRUE))
  plot(lg_vec,type="l",main="Number of Stocks in Long Portfolio")
  
  # Subset returns dataframe to only have returns for stocks whose betas were calculated
  # First row is removed to match dimension of weight dataframes
  stks <- df_all[,in_in][-1,] 
  t_endpoints <- endpoints(stks, x)
  stks <- stks[t_endpoints,]
  stks <- stks['2002-02/2017-06']
  stks[is.na(stks)] <- 0 # Convert NAs to 0s

  # Pret_long is the portfolio return of the stocks that were bought
  # Pret_short is the portfolio return of the stocks that were sold short
  pret_long <- coredata(long_wt) * coredata(stks)
  pret_short <- coredata(short_wt) * coredata(stks)
  
  # P is the sum (really weighted sum given returns were multiplied by weights above) of all the returns of stocks that were bought on a weekly 
  # Same is true for S for the stocks that were sold short on a weekly basis
  p <- apply(pret_long,1,sum) 
  s <- apply(pret_short,1,sum)
  pt <- cbind.data.frame(p,s) # combine both p & s into a dataframe
  
  if (x == "years"){
    idx <- seq(as.Date("2003-01-01"),as.Date("2017-01-01"),by = x)
  } else {
    idx <- seq(as.Date("2002-02-07"),as.Date("2017-06-30"),by = x)
  }
  pt <- as.xts(pt,order.by = idx) # convert to xts object
  # calculate total portfolio return if long and short portfolios were equal weighted
  f <- p + s
  f.ts <- as.xts(f,order.by = idx) # convert total portfolio return to xts object
  
# Cumulative return is then calculated and plotted. Anything below the horizontal line means the investment lost money. Other performance statistics are then plotted/printed using functions from the PerformanceAnalytics package.
  cumret <- cumprod(1+f.ts)
  plot.xts(cumret)
  abline(h=1)
  charts.PerformanceSummary(f.ts)
  print(Return.annualized(f.ts))
  print(SharpeRatio.annualized(f.ts))
}
```

### Test Results for Weekly Rebalance:

```{r w-rebal}
bab_test("week")
```

### Test Results for Monthly Rebalance:

```{r m-rebal}
bab_test("months")
```

### Test Results for Quarterly Rebalance:

```{r q-rebal}
bab_test("quarter")
```


### Test Results for Annual Rebalance:

```{r a-rebal}
bab_test("years")
```

## CONCLUSION

**Main Conclusion:**

Unlike the authors of the research paper referenced in the beginning, I wasn't able to find any significant advantage from an investor that shorts high-beta stocks and buys long-beta stocks.

**Limitations:**

The discrepancy in these results versus the referenced research paper could be due to various reasons such as:
1. This dataset isn't as exhaustive and suffers from survivorship basis.
2. Calculation of beta on a daily frequency was too onerous for the purpose of this project, which differs from the approach taken in the research paper.
